---
title: "Red Wood Data Analysis Report"
author: "Andrew, Dylan (Shi-Ting Lu)"
output:
  pdf_document: default
  html_document: default
---
## Set Up
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#library(tidyverse)
#library(ggplot2)
#library(caret)
```

```{r, read-in-data}
all <- read.csv("sonoma-data-all.csv")
log <- read.csv("sonoma-data-log.csv")
net <- read.csv("sonoma-data-net.csv")
dates <- read.csv("sonoma-data-net.csv")
```


## 1 Data collection

#### (a - Paper Summary)
The purpose of the study is twofold: capture information about a single redwood tree canopy over time and provide a roadmap for future macroscopic studies using a multi-sensor network.

The data was collected in Sonoma, California on a single Redwood tree over a period of days at consistent time intervals during the late spring/early summer.

Based on this study, researchers were able to verify the existence of dynamic spatio-temporal gradients surrounding the tree and prove that complex biological theories can be validated using this measurement framework. Researchers highlighted lesson's learned, beneficial for future studies, highlighting sensor sensitivity based on positioning and yield issues from memory/network constraints.

#### (b - Data Collection)
# How are sensors deployed?
Nodes (sensor housing) were attached to the body of one redwood tree at various radial, angular, and vertical distances. At  roughly 2-meter spacing intervals the first sensor was placed 15m from ground level and the last sensor at 70m above ground level. The majority of nodes used in the analysis were placed on the west side of the tree about 0.1m - 1m from the tree trunk. Several nodes were placed outside of the measurement envelope to monitor readings in the immediate vicinity.

Nodes had two means of capturing/transferring data: on-site logger stored readings on-device and a separate workflow transferred readings over-the-wire to gateway (make sure this is correct when looking at file).

Sensors were calibrated before being deployed in the field using two trials called roof and chamber. In the roof exercise, nodes were placed in direct view of sunlight atop a building to test the PAR measurements and which were compared to a well known reference. In the chamber phase, temperature and humidity sensors exposed to wide range of conditions: from 5-30 degrees Celsius and between 20-90 %RH. Before being deployed the data harvesting querying was tested in field on a sample tree in similar conditions to verify communication between nodes and on-site internet connected gateway.

# What is the duration of data recording?
Data was recorded over period of almost 44 days from 4/27/2004 5:10pm to 6/10/2004 2:00pm. Measurements were taken every 5 minutes and battery operated nodes were duty cycled to conserve power when not operating (on for 4 seconds to take measurements then turned off until next reading). 

# What are the main variables of interest
Researchers were interested in traditional climate variables temperature, humidity, and light levels.	Photosynthetically Active Radiation (PAR) represented light levels and included light wavelengths between 350nm–700nm

# What is difference between data in two different files?


## 2 Data cleaning

#### (a) (Dylan)

```{r hist-of-variables}
print("Hello World!")
```


#### (b) (Dylan)

#### (c) (Dylan)

#### (d) (Dylan)

#### (e) (Dylan)

***

#### (a) (Andrew)
```{r}

```

#### (b) (Andrew)

#### (c) (Andrew)

#### (d) (Andrew)

#### (e) (Andrew)

***

## 3 Data Exploration

#### (a) (Dylan)

#### (b) (Dylan)

#### (c) (Dylan)

#### (d) (Dylan)

***

#### (a) (Andrew)

#### (b) (Andrew)

#### (c) (Andrew)

#### (d) (Andrew)

```{r, Andrew}
head(all)
```